Flabby and injective objects in toposes

### 2018-10-21 paper-flabby-objects.tex
First approximation to an introduction

XXX: introduction
% XXX: In Einleitung interne Sprache erinnern.

>>>>

As is nowadays well-established, any topos supports an \emph{internal language}
which can be used to reason about the objects and morphisms of the topos in a
naive element-based language, allowing us to pretend that the objects are plain
sets (or types) and the morphisms are plain maps between those sets. The
internal language is sound with respect to intuitionistic reasoning, whereby
any intuitionistic theorem holds in any topos.

The internal language of a sheaf topos enables \emph{relativization by
internalization}. For instance, by interpreting the
proposition \begin{quote}``in any short exact sequence of modules, if the two
outer ones are finitely generated then so is the middle one''\end{quote} of
intuitionistic commutative algebra internally to the topos of sheaves over a
space~$X$, we obtain the geometric analogue \begin{quote}``in any short exact
sequence of sheaves of modules over~$X$, if the two outer ones are of finite
type then so is the middle one''.\end{quote}
This way of deducing geometric theorems provides conceptual clarity, reduces
technical overhead and justifies certain kinds of ``fast and loose reasoning''
typical of informal algebraic geometry. As soon as we go beyond the fragment of
geometric sequents and consider more involved first-order or even higher-order
statements, also significant improvements in proof length and proof complexity
can be gained. For instance, Grothendieck's generic freeness lemma admits a
short and simple proof in this framework, while previously-published proofs
proceed in a somewhat involved series of reduction steps and require a fair
amount of prerequisites in commutative
algebra~\cite{blechschmidt:phd,blechschmidt:generic-freeness}.

The practicality of this approach hinges on the extent to which the dictionary
between internal and external notions has been worked out. For instance, the
simple example displayed above hinges on the dictionary entry stating that a
sheaf of modules is of finite type if and only if it looks like a
finitely-generated module from the internal point of view. The motivation for
this note was to find internal characterizations of flabbiness and of higher
direct images, and the resulting entries are laid out in
Section~\ref{sect:flabby-objects} and in
Section~\ref{sect:higher-direct-images}: A sheaf is flabby if and only if, from
the internal point of view, it is a flabby set -- a notion introduced in
Section~\ref{sect:flabby-sets} below; and higher direct images look like sheaf
cohomology from the internal point view.

As a byproduct, we demonstrate how the notion of flabby sets is a useful
organizing principle in the study of injective objects. We employ flabby sets
to give a new proof of Roswitha Harting's result that injectivity of sheaves is
a local notion~\cite{harting:remark} and that a sheaf is injective if and only
if it is injective from the internal point of
view~\cite{harting:locally-injective}, which she stated (in slightly different
language) for sheaves of abelian groups. We use the opportunity to correct a
small mistake of her, namely claiming that no generalization to sheaves of
modules is possible.

A further motivation for this note was our desire to seek a constructive
account of sheaf cohomology. Sheaf cohomology is commonly defined using
injective resolutions, which can fail to exist in the absence of the axiom of
choice~\cite{blass:inj-proj-axc}, but it is also known that flabby resolutions
can be used in their stead, making them the obvious candidate for a
constructively sensible replacement of the usual definition. However, we show
in Section~\ref{sect:in-eff} and in Section~\ref{sect:conclusion} that flabby
resolutions present their own challenges, and in summary we failed to reach our
goal. To the best of our knowledge, the problem of giving a constructive accout
of sheaf cohomology is still open.

In view of almost 80 years of sheaf cohomology, this state of affairs is
slightly embarassing, challenging the call that ``once [a] subject is better
understood, we can hope to refine its definitions and proofs so as to
avoid that axiom''~\cite[Section~3.4]{hott}.

A constructive account of sheaf cohomology would be highly desirable, not only
out of a philosophical desire to obtain a deeper understanding of the
foundations of sheaf cohomology, but also to: use the tools of sheaf cohomology
in the internal setting of toposes, thereby extending the applicability of
relativization by internalization; and to carry out \emph{integrated
developments} of algorithms for computing sheaf cohomology, where we would
extract algorithms together with termination and correctness proofs from a
hypothetical constructive account.
